Memes: an R interface to the MEME Suite

### 2020-10-22 manuscript.Rmd
More edits.

Biopolymers, such as DNA and protein, perform varying functions based on their
primary sequence. Short, repeated sequences, or "motifs" represent functional
units within biopolymers which can act as interaction surfaces, create
structure, or have enzymatic activity. Identification of shared motifs across
multiple sequences can provide evidence for shared function, thus the ability to
identify, classify, and compare motifs represents a key step in the analysis of
biological sequences. Although tools that implement these features have
historically existed in the Bioconductor landscape, they often rely on
proprietary data types with poor interoperability with other tools. Furthermore,
several of these packages will be deprecated in the next Bioconductor release,
leaving a gap in functionality with no clear alternatives.

The MEME Suite is a series of commandline and web-based tools that provide a variety
of features for motif analysis such as *de-novo* motif discovery, enrichment
testing, comparison, and matching \cite{meme-suite}. The MEME Suite is a widely
utilized set of tools, with over 22,000 citations to date, and greater than 30,000 unique
users of the webserver implementation annually \cite{meme-usage-report-page,
google-scholar-meme}.

The MEME Suite family of tools provides a comprehensive set of utilities to
interrogate motif content across a broad range of biological contexts. However,
several factors limit the full potential of these tools for use in data
analysis. MEME Suite tools require carefully formatted inputs to achieve full
functionality, yet few tools exist to simplify the process of data formatting,
requiring instead that users write custom code to prepare their data.
Furthermore, the output data from each MEME Suite tool often have complex
structures which must be parsed to extract the full suite of information, again
requiring users to write custom code for this task. Finally, although the
data-generation capabilities of the MEME Suite are excellent, the tools lack
powerful ways to visualize the results. These factors act as barriers to
adoption and preclude deep analysis of the data.

Here we present `memes`, a motif analysis package which provides a seamless
interface to MEME Suite tools within R. `memes` uses base R and Bioconductor
data types for data input and output, facilitating better integration with
common analysis tools like the `tidyverse` and other Bioconductor packages.
Unlike the commandline implementation, `memes` outputs also function as inputs
to other MEME Suite tools, allowing simple construction of motif analysis
pipelines without additional data processing steps. Additionally, R/Bioconductor
data structures provide a full-featured representation of MEME Suite output
data, providing users quick access to all relevant data structures with simple
syntax.

In addition to wrapping MEME Suite utilities, `memes` extends the functionality
of `universalmotif` to provide base R and `tidyverse` compatible syntax for
manipulating and analyzing motif metadata. These features allow motif databases
to integrate external data to aid analysis, such as filtering a set of
transcription factor motifs based on gene expression in the tissue of interest.

@@

# Results 
## Design & Implementation

@@

MEME Suite tools are run on the commandline and use files stored on-disk as
input while returning a series of output files containing varying data types.
`memes` reduces each MEME Suite tool to an R function which internally handles
file I/O such that all MEME Suite calling functions accept R objects as input,
and return R objects as output. Sequence inputs are stored in `BioStrings`
format, an R/Bioconductor package for storing biopolymer sequence data
\cite{BioStrings}. Motif inputs are passed as `universalmotif` objects, another
R/Bioconductor package for representing motif matrices and their associated
metadata \cite{universalmotif}. `memes` will also always accept paths to files
on disk, such as fasta files for sequence inputs, and meme format files for
motif inputs, reducing the need to read large files into memory to run MEME
Suite tools.

Although MEME Suite tools accept consistent data type inputs, they do not return
consistent data structures across tools. For example, the tool TomTom, which
compares query motifs to a database of known motifs, returns tab-separated
results which do not contain the frequency matrix representations of the matched
motifs. Instead, users must write custom code to parse these matrices back out
from the input databases, creating additional friction for creating novel data
visualizations and exploratory analyses. Often, output data are also returned in
XML format, which although they hold all relevant information, are difficult to
parse. To our knowledge, no publicly available tools exist to parse MEME Suite
XML formatted data, therefore users must write custom code to extract these
data. `memes` exports data import functions for each supported MEME Suite tool,
which import these data as R data.frames. These functions also work internally to
create the seamless file I/O when calling MEME Suite tools, but the import
functions can be called directly by users to import previously generated data
for use in R.

@@

Many of the MEME Suite tools, such as MEME, DREME, or TomTom return data which
describe properties of motifs. For example, MEME and DREME return de-novo
discovered motifs from query sequences along with statistical information about
their enrichment. The `universalmotif` data structure is a powerful
R/Bioconductor representation for motif matrices and their associated metadata \cite{universalmotif}. 
However, `universalmotif` objects are not ideal for performing bulk manipulation
or queries of metadata, nor do they support complex data structures such as
nested elements. Many MEME Suite analysis pipelines involve identifying motifs
which are further analyzed by multiple tools, creating additional metadata which
cannot be easily tracked in bulk using `universalmotif` format. Thus
an ideal representation for these data is one which can hold an unlimited amount
of metadata, contain complex hierarchies, and be easily manipulated using
standard analysis tools. `memes` therefore extends the `universalmotif` format
by providing data.frames holding motif metadata which retain links to their
`universalmotif` representation. These universalmotif
data.frames combine the advantages of the `universalmotif` framework for motif
manipulation with the extensibility and flexibility of R data.frames. These
`universalmotif` data.frames form the basis for a majority of `memes` data
outputs. These structures are also valid input types to `memes` functions, and
when used as such, output data are appended as new columns to the input data,
ensuring data provenance.

@@

Motif analysis is important for assays like ChIP-seq which store data as genomic
coordinates, not sequence. This is in conflict with MEME Suite which is designed
to study sequences. Therefore, users must write custom code to extract DNA
sequence of their ranges of interest. Although tools like `bedtools getfasta`
can simplify the process of extracting DNA sequence from genomic coordinates,
some MEME tools require fasta headers to be specially formatted to extract full
functionality, requiring users to write additional custom code to properly
format input data.

`memes` provides the `get_sequence` function which automates the extraction of
DNA sequence from genomic coordinates while simultaneously producing MEME Suite
formatted fasta headers. `get_sequence` accepts genomic-range based inputs in
`GenomicRanges` format, the *de-facto* standard for genomic coordinate
representation in R \cite{genomicranges}. Other common genomic coordinate
representations, such as bed format, are easily imported as `GenomicRanges`
objects into R using preexisting import functions, meaning users must not write
any custom import functions to work with range-based data using `memes`. 
Sequences are returned in `Biostrings` format and can therefore be used as
inputs to all `memes` commands or as inputs to other R/Bioconductor functions
for sequence analysis.

During analysis of multiomic data, users often uncover different functional
categories of genomic regions through integration of multiple datasets. In these
instances, motif analysis acts as a powerful tool to examine whether differences
in motif content is associated with membership of distinct categories. In some
cases, performing differential motif analysis (using one category as the "test"
set and another as the "control") can uncover biologically relevant motifs which
are enriched in one category relative to another. Although the MEME Suite allows
differential enrichment testing, it does not inherently provide a mechanism for
analyzing groups of sequences in parallel, or performing motif analysis with any
understanding of data categories. The `memes` framework enables "data-aware"
motif analysis workflows by allowing named lists of `Biostrings` objects as
input to each function. If using `GenomicRanges`, users can split peak data on a
metadata column using the base R `split` function, and use this result as input
to `get_sequence`, which will produce a list of `BioStrings` objects where each
entry is named after each data category from the split column. When a list is
used as input to a `memes` function, it runs the corresponding MEME Suite tool
for each object in the list. Users can also pass the name(s) of a category to
the `control` argument to automatically enable differential analysis of all
other list members against the control category sequences.  Therefore, `memes`
enables complex data-aware differential motif analysis workflows using simple
syntax to extend the capabilities of the MEME Suite.

### Table of supported MEME Suite Tools
 - comparison to other tools?
 - Maybe only do this if asked

## Data visualization
The MEME Suite provides a small set of data visualizations which have limited
customizability. `memes` leverages the advantages of the R graphics environment
to provide novel data visualizations which are infinitely customizable by the user.

The TomTom tool allows users to compare unknown motifs to a set of known motifs
to identify a best match. Visual inspection of motif comparison data is a key
step in assessing the quality and correctness of a match. The `view_tomtom_hits`
function allows users to compare query motifs with the list of potential matches
as assigned by TomTom. Often, matching tools are useful for reducing the
search-space for potential candidates, but users are free to leverage their
understanding of the biology to override these matches. The `force_best_match`
function allows users to quickly reassign the TomTom best match to a
lower-ranked, but still statistically significant match to ease downstream data
processing.

The AME tool searches for enrichment of motifs within groups of sequences
relative to a control set. The meaningful result from this tool is the
statistical parameter (for example, a p-value) associated with the significance
of motif enrichment. However, AME does not provide a mechanism for visualizing
these results. Additionally, AME is often used to examine motif content of
multiple groups of sequences, therefore a plot which can represent statistical
significance and group membership of a motif match could be a useful way of
visualizing the data. `memes` visualized AME results with the `ame_plot_heatmap`
function, which returns a `ggplot2` formatted heatmap comparing motif
significance within one or multiple groups. Several options exist to customize
the heatmap values to help capture different aspects of the output data. The
`ame_compare_heatmap_methods()` function enables users to compare the
distribution of values between samples in order to select a threshold which
accurately captures the dynamic range of their results.

## Containerized analysis maximizes availability and facilitates reproducibility
`memes` relies on a locally installed version of the MEME Suite which is
accessed by the users R process. Although R is available on Windows, Mac, and
Linux operating systems, the MEME Suite is incompatible with Windows, limiting
its adoption by Windows users. Additional barriers also exist to installing a
local copy of the MEME Suite on compatible systems. The MEME Suite relies on
several system-level dependencies whose installation can be difficult for novice
users. Finally, some tools in the MEME Suite use python to generate shuffled
control sequences for analysis, which presents a reproducibility issue as the
random number generation algorithm changed from python2.7 to python3. The MEME
Suite will build and install on both python2.7 and python3 systems, therefore
without careful consideration of this behavior, the same code run on two systems
may not produce identical results, even if using the same major version of the
MEME Suite. In order to democratize access to the MEME Suite on unsupported
operating systems, and to facilitate reproducible motif analysis using `memes`, 
we have also developed docker and singularity (**NOTE:** singularity is broken
for `rocker/rstudio`, so maybe omit this until fixed) containers containing a
preinstalled version of the MEME Suite along with an R/Bioconductor analysis
environment including the most recent version of `memes` and its dependencies.
As new container versions are released, they are version tagged and stored to
ensure reproducibility while allowing updates to the container.

@@

Here we briefly highlight each of `memes` current features for analyzing
ChIP-seq data. Additional detailed walkthroughs for each supported MEME Suite
tool, and a complete worked example using `memes` to analyze ChIP-seq data can be found
in the `memes` vignette and the package website
([snystrom.github.io/memes](snystrom.github.io/memes)). In the following
example, we reanalyze recent work examining the causal relationship between the
binding of the transcription factor E93 to changes in chromatin accessibility
during *Drosophila* wing development \cite{Nystrom, Niederhuber 2020}. Here, we
utilize ChIP-seq peak calls for E93 which have been annotated according to the
change in chromatin accessibility observed before & after E93 binding. These
data are an idiomatic example of range-based genomic data (E93 ChIP peaks)
containing additional groupings (the chromatin accessibility response following
DNA binding) whose membership may be influenced by differential motif content.
We show how `memes` syntax allows complex analysis designs, how `memes`
utilities enable deep interrogation of results, and how `memes` flexible data
structures empower users to integrate the `memes` workflow with tools offered by
other R/Bioconductor packages.

>>>>

Biopolymers, such as DNA and protein, perform varying functions based on their
primary sequence. Short, repeated sequences, or "motifs" represent functional
units within biopolymers which can act as interaction surfaces, create
structure, or have enzymatic activity. Identification of shared motifs across
multiple sequences can provide evidence for shared function, for example,
identification of kinase substrates based on similarities in phosphorylation
site sequence, or characterizing DNA elements based on shared transcription
factor binding sequences (TODO: ref kinase motif:
https://doi.org/10.1016/0968-0004(90)90073-K). Thus the ability to
identify, classify, and compare motifs represents a key step in the analysis of
biological sequences. **TODO: R/Bioconductor transition sentence**. Although
tools that implement these features have historically existed in the
Bioconductor landscape, they often rely on proprietary data types with poor
interoperability with other tools, creating barriers to comprehensive motif
analysis within R. 
**TODO: fix this section, how to transition?**
Furthermore, popular motif analysis packages, MotIV, and motifRG 
will be deprecated in the next Bioconductor release,
leaving a gap in functionality with no clear alternatives.

The MEME Suite is a series of commandline and web-based tools that provide a variety
of features for motif analysis such as *de-novo* motif discovery, enrichment
testing, comparison, and matching \cite{meme-suite}. The MEME Suite is a widely
utilized set of tools, with over 22,000 citations to date, and greater than 30,000 unique
users of the webserver implementation annually \cite{meme-usage-report-page,
google-scholar-meme}. The MEME Suite family of tools provides a comprehensive
set of utilities to interrogate motif content across a broad range of biological
contexts. However, several factors limit the full potential of these tools for
use in data analysis. MEME Suite tools require carefully formatted inputs to
achieve full functionality, yet few tools exist to simplify the process of data
formatting, requiring instead that users write custom code to prepare their
data, or prepare the inputs by hand, both of which have the potential to be
error prone without rigorous testing. Furthermore, the output data from each
MEME Suite tool often have complex structures that must be parsed to extract the
full suite of information, again requiring users to write custom code for this
task. Finally, although the data-generation capabilities of the MEME Suite are
excellent, the tools lack powerful ways to visualize the results. Collectively,
these factors act as barriers to adoption and preclude deep analysis of the
data, as well as communication of results to the scientific community.

Here we present `memes`, a motif analysis package that provides a seamless
interface to MEME Suite tools within R. `memes` uses base R and Bioconductor
data types for data input and output, facilitating better integration with
common analysis tools like the `tidyverse` and other Bioconductor packages.
Unlike the commandline implementation, `memes` outputs also function as inputs
to other MEME Suite tools, allowing simple construction of motif analysis
pipelines without additional data processing steps. Additionally, R/Bioconductor
data structures provide a full-featured representation of MEME Suite output
data, providing users quick access to all relevant data structures with simple
syntax. **TODO: Depending on length restrictions, end here**

In addition to wrapping MEME Suite utilities, `memes` extends the functionality
of `universalmotif` to provide base R and `tidyverse` compatible syntax for
manipulating and analyzing motif metadata. These features allow users to
integrate external data with motif databases to aid analysis, such as filtering
a set of transcription factor motifs based on gene expression in the tissue of
interest, then using that filtered set to search for motifs associated with
expressed factors.

@@

## Design & Implementation

@@

MEME Suite tools are run on the commandline and use files stored on-disk as
input while returning a series of output files containing varying data types.
`memes` reduces each MEME Suite tool to an R function which internally handles
file I/O such that all MEME Suite calling functions accept R objects as input,
and return R objects as output. Sequence inputs are stored in `BioStrings`
format, an R/Bioconductor package for storing biopolymer sequence data
\cite{BioStrings}. Motif inputs are passed as `universalmotif` objects, another
R/Bioconductor package for representing motif matrices and their associated
metadata \cite{universalmotif}. `memes` will also accept paths to files
on disk, such as fasta files for sequence inputs, and meme format files for
motif inputs, reducing the need to read large files into memory.

**Restructure below:**
 - MEME tools return HTML reports for each tool which display data in user-friendly way, but poor for data processing. However, some of the simple raw data (like tsv data) don't contain the full set of statistics reported to the user. These data are contained in XML files, which are difficult to parse.

Although MEME Suite tools accept consistent data type inputs, they do not return
consistent data structures across tools. For example, the TomTom tool, which
compares query motifs to a database of known motifs, returns tab-separated
results that do not contain the frequency matrix representations of the matched
motifs. Instead, users must write custom code to parse these matrices back out
from the input databases, creating additional friction for creating novel data
visualizations and exploratory analyses. Often, output data are also returned in
XML format, which although they hold all relevant information, are difficult to
parse. To our knowledge, no publicly available tools exist to parse MEME Suite
XML formatted data, therefore users must write custom code to extract these
data. `memes` exports data import functions for each supported MEME Suite tool,
which import these data as R data.frames. These functions also work internally to
create the seamless file I/O when calling MEME Suite tools, but the import
functions can be called directly by users to import previously generated data
for use in R.

@@

Many of the MEME Suite tools return files which
contianing descriptive properties of motifs. For example, MEME and DREME return de-novo
discovered motifs from query sequences along with statistical information about
their enrichment. The `universalmotif` data structure is a powerful
R/Bioconductor representation for motif matrices and their associated metadata \cite{universalmotif}. 
However, `universalmotif` objects are not ideal for performing bulk manipulation
or queries of metadata, nor do they support complex data structures such as
nested elements. Many MEME Suite analysis pipelines involve identifying motifs
which are further analyzed by multiple tools, creating additional metadata which
cannot be easily tracked in bulk using `universalmotif` format. Thus
an ideal representation for these data is one which can hold an unlimited amount
of metadata, contain complex hierarchies, and be easily manipulated using
standard analysis tools. `memes` therefore extends the `universalmotif` format
by providing data.frames holding motif metadata which retain links to their
`universalmotif` representation. These universalmotif
data.frames combine the advantages of the `universalmotif` framework for motif
manipulation with the extensibility and flexibility of R data.frames. These
`universalmotif` data.frames form the basis for a majority of `memes` data
outputs. These structures are also valid input types to `memes` functions, and
when used as such, output data are appended as new columns to the input data,
ensuring data provenance.

@@

Motif analysis is often employed in ChIP-seq analysis, in which data is stored as genomic
coordinates rather than sequence. However, MEME Suite tools are designed to work
with sequences. While existing tools such as bedtools can extract DNA sequence
from gneomic coordinates, some MEME tools require fasta headers to be
specifically formatted. As a result, users must write custom code to extracd DNA
sequence for their genomic ranges of interest.

The `memes` function `get_sequence` automates extraction of
DNA sequence from genomic coordinates while simultaneously producing MEME Suite
formatted fasta headers. `get_sequence` accepts genomic-range based inputs in
`GenomicRanges` format, the *de-facto* standard for genomic coordinate
representation in R \cite{genomicranges}. Other common genomic coordinate
representations, such as bed format, are easily imported as `GenomicRanges`
objects into R using preexisting import functions, meaning `memes` users do not
have to write any custom import functions to work with range-based data using
`memes`.  Sequences are returned in `Biostrings` format and can therefore be
used as inputs to all `memes` commands or as inputs to other R/Bioconductor
functions for sequence analysis.

During analysis of multiomic data, users can identify different functional
categories of genomic regions through integration with orthogonal datasets. **TODO
For example, .**

In these instances, motif analysis acts as a powerful tool to
examine whether differences in motif content is associated with membership of
distinct categories. In some cases, performing differential motif analysis
(using one category as the "test" set and another as the "control") can uncover
biologically relevant motifs which are enriched in one category relative to
another. Although the MEME Suite allows differential enrichment testing, it does
not inherently provide a mechanism for analyzing groups of sequences in
parallel, or performing motif analysis with any understanding of data
categories. The `memes` framework enables "data-aware" motif analysis workflows
by allowing named lists of `Biostrings` objects as input to each function. If
using `GenomicRanges`, users can split peak data on a metadata column using the
base R `split` function, and then use this result as input to `get_sequence`, which
will produce a list of `BioStrings` objects where each entry is named after the
data categories from the split column. When a list is used as input to a `memes`
function, it runs the corresponding MEME Suite tool for each object in the list.
Users can also pass the name(s) of a category to the `control` argument to
enable differential analysis of the remaining list members against the
control category sequences. Therefore, `memes` enables complex data-aware
differential motif analysis workflows using simple syntax to extend the
capabilities of the MEME Suite.

## Data visualization
The MEME Suite provides a small set of data visualizations that have limited
customizability. `memes` leverages the advantages of the R graphics environment
to provide a wide range of data visualization options that are highly
customizable. We describe two scenarios below.

The TomTom tool allows users to compare unknown motifs to a set of known motifs
to identify the best match. Visual inspection of motif comparison data is a key
step in assessing the quality and accuracy of a match. The `view_tomtom_hits`
function allows users to compare query motifs with the list of potential matches
as assigned by TomTom. The `force_best_match` function allows users to
reassign the TomTom best match to a lower-ranked (but still statistically
significant) match in order to highlight motifs with greater biological
relevance (e.g. to skip over a transcription factor that is not expressed in the
experimental sample, or when two motifs are matched equally well).

The AME tool searches for enrichment of motifs within a set of experimental
sequences relative to a control set. The meaningful result from this tool is the
statistical parameter (for example, a p-value) associated with the significance
of motif enrichment. However, AME does not provide a mechanism for visualizing
these results.

The `ame_plot_heatmap` function in `memes` returns a `ggplot2` formatted heatmap of
statistical significance of motif enrichment. If AME is used to examine motif
content of multiple groups of sequences, the `ame_plot_heatmap` function can also
return a plot comparing motif significance within multiple groups. Several
options exist to customize the heatmap values in order to capture different aspects
of the output data. The `ame_compare_heatmap_methods()` function enables users
to compare the distribution of values between samples in order to select a
threshold that accurately captures the dynamic range of their results.

## Containerized analysis maximizes availability and facilitates reproducibility
`memes` relies on a locally installed version of the MEME Suite which is
accessed by the user's R process. Although R is available on Windows, Mac, and
Linux operating systems, the MEME Suite is incompatible with Windows, limiting
its adoption by Windows users. Additional barriers also exist to installing a
local copy of the MEME Suite on compatible systems, for example, the MEME Suite
relies on several system-level dependencies whose installation can be difficult
for novice users. Finally, some tools in the MEME Suite use python to generate
shuffled control sequences for analysis, which presents a reproducibility issue
as the random number generation algorithm changed between python2.7 and python3. The
MEME Suite will build and install on both python2.7 and python3 systems,
therefore without careful consideration of this behavior, the same code run on
two systems may not produce identical results, even if using the same major
version of the MEME Suite. In order to increase access to the MEME Suite on
unsupported operating systems, and to facilitate reproducible motif analysis
using `memes`, we have also developed docker and singularity (**TODO: NOTE:**
singularity is broken for `rocker/rstudio`, so maybe omit this until fixed)
containers containing a preinstalled version of the MEME Suite along with an
R/Bioconductor analysis environment including the most recent version of `memes`
and its dependencies.  As new container versions are released, they are
version-tagged and stored to ensure reproducibility while allowing updates to
the container.

@@

Here we briefly highlight each of `memes` current features for analyzing
ChIP-seq data. Additional detailed walkthroughs for each supported MEME Suite
tool, and a complete worked example using `memes` to analyze ChIP-seq data can be found
in the `memes` vignettes and the package website
([snystrom.github.io/memes](snystrom.github.io/memes)). In the following
example, we reanalyze recent work examining the causal relationship between the
binding of the transcription factor E93 to changes in chromatin accessibility
during *Drosophila* wing development \cite{Nystrom, Niederhuber 2020}. Here, we
utilize ChIP-seq peak calls for E93 which have been annotated according to the
change in chromatin accessibility observed before & after E93 binding. These
data are an idiomatic example of range-based genomic data (E93 ChIP peaks)
containing additional groupings (the chromatin accessibility response following
DNA binding) whose membership may be influenced by differential motif content.
We show how `memes` syntax allows complex analysis designs, how `memes`
utilities enable deep interrogation of results, and how `memes` flexible data
structures empower users to integrate the `memes` workflow with tools offered by
other R/Bioconductor packages.
